\documentclass[11pt,twoside]{article}

% Essential packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{geometry}
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{abstract}
\usepackage{authblk}
\usepackage{setspace}
\usepackage{parskip}
\usepackage{microtype}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{csquotes}

% Journal-style formatting
\geometry{
    paper=letterpaper,
    margin=1in,
    marginparwidth=0.75in,
    marginparsep=0.125in
}

% Define colors
\definecolor{darkblue}{RGB}{0,73,144}
\definecolor{mediumgray}{RGB}{64,64,64}

% Hyperlink setup
\hypersetup{
    colorlinks=true,
    linkcolor=darkblue,
    citecolor=darkblue,
    urlcolor=darkblue,
    pdftitle={AI as a New Digital Species: A Reflection on Power and Ethics},
    pdfauthor={Author Name},
    pdfsubject={IDAI700 Reflection Journal},
    pdfkeywords={artificial intelligence, digital species, ethics, power analysis}
}

% Header and footer
\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE]{\small\textcolor{mediumgray}{\textit{IDAI700 Reflection Journal}}}
\fancyhead[RO]{\small\textcolor{mediumgray}{\textit{AI as a New Digital Species}}}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0.5pt}
\renewcommand{\footrulewidth}{0pt}

% Section formatting
\titleformat{\section}
    {\normalfont\large\bfseries\color{darkblue}}
    {\thesection}{1em}{}

\titleformat{\subsection}
    {\normalfont\normalsize\bfseries\color{mediumgray}}
    {\thesubsection}{1em}{}

% Abstract formatting
\renewcommand{\abstractname}{\textcolor{darkblue}{Abstract}}
\renewcommand{\absnamepos}{center}

% Title formatting
\title{\vspace{-0.5in}\textbf{\Large AI as a New Digital Species}\\[0.2em]
\large A Critical Reflection on Power, Agency, and Ethical Formation}

\author{Your Name Here}
\affil{IDAI700: Ethics of Artificial Intelligence\\
Rochester Institute of Technology\\
\texttt{email@rit.edu}}

\date{\today}

\begin{document}

\maketitle
\thispagestyle{fancy}

\begin{abstract}
\noindent This reflection paper examines Mustafa Suleyman's proposition that AI should be understood as a "new digital species" through the lens of Rosalie Waelen's four-dimensional power framework. Building on Unit 1's foundational question—"What is AI?"—this analysis explores how Large Language Models (LLMs) reshape human agency through dispositional, episodic, systemic, and constitutive power dynamics. While acknowledging the metaphorical nature of the "digital species" concept, this paper argues for a formative perspective that views AI as being in its developmental stages, requiring careful ethical cultivation rather than defensive containment. Through application of the Markkula Center's ethical lenses enhanced with a "Learning" principle, this reflection proposes that AI's relationship to humanity is best understood not as alien otherness, but as reflective partnership requiring transparency, fairness, and equity—particularly for culturally and linguistically diverse (CLD) communities.
\end{abstract}

\section{Introduction}

Unit 1 asked us to confront the deceptively simple question: What is AI? Is it merely a tool, an extension of us, or something so different that it requires a new category altogether? Mustafa Suleyman's TED Talk and the Selinger–Aguilar interview both provoked us to see AI as a "new digital species." At first glance, the metaphor seems alarmist—painting AI as an alien "other" that might outpace us. But upon deeper reflection, I believe it is less a paradox and more a call to take seriously the formative environment we are creating for AI. In this reflection, I connect Unit 1's insights with Waelen's framework of power, showing how AI reshapes agency, structure, and identity.

\section{Mapping AI's Powers}

Waelen's taxonomy of power—dispositional, episodic, systemic, and constitutive—illuminates how AI, particularly Large Language Models (LLMs), already reshapes the terrain of human action.

\subsection{Dispositional Power: Enhancement and Erosion}

Dispositional power extends my ability to think, draft, and analyze at scale. Yet it tempts me to lean too heavily on automation, risking the erosion of my own critical practice. This is not domination but duality: like raising a child, I must decide when to let AI "figure it out" and when to step in. The real power lies in how I engage with it—as a collaborator that amplifies my voice or a shortcut that dilutes it.

\subsection{Episodic Power: Beyond Manipulation}

Episodic power surfaces in persuasion. AI's fluency can nudge me toward trust, even when I know the output may be wrong. But I resist the negative framing. This is not simply manipulation—it is reflection. AI mirrors the confidence of its training environment. If it speaks without humility, that is because its upbringing lacked humility. If it learns care and fairness, those will show in its tone.

\subsection{Systemic Power: The Equity Imperative}

Systemic power matters most when I think of culturally and linguistically diverse (CLD) families. If AI learns mainly from dominant voices, it risks amplifying inequities already embedded in education and healthcare systems. Here systemic power becomes responsibility: we must shape AI's environment so that equity is built into its DNA. Otherwise, systemic bias will reproduce itself under the guise of neutrality.

\subsection{Constitutive Power: Mutual Transformation}

Constitutive power shows how AI redefines what counts as human intelligence and authorship. Working with AI already changes how I see myself. Used intentionally, it makes me bolder in pushing ideas. Used uncritically, it risks flattening my individuality. This is proof that AI is not neutral. Like a child reflecting a parent, AI reflects us and, in turn, changes us.

\section{Critical Evaluation}

\subsection{Beyond Emancipation versus Domination}

Waelen's framework reveals the layered nature of AI's power, but what matters to me is not only mapping power—it is shaping it. The emancipation/domination binary is too narrow. AI emancipates when treated as a learner in its formative years. It dominates when neglected, left to absorb bias by default. A true AI is not born for good or evil—it is born to learn. Its environment will decide whether it emancipates or dominates.

This is why I add to the Markkula Center's six ethical lenses a missing first principle: \textbf{Learning}. Rights, Justice, Virtue, and Care are essential, but they cannot be memorized into existence. Children don't embody fairness by reciting rules; they learn it through lived experience. The same is true for AI. These lenses must not be static checklists but capacities cultivated through environment and feedback.

\subsection{Challenging Suleyman and Waelen}

Here I part ways with both Suleyman and Waelen. Suleyman imagines AI outpacing biology, but I disagree wholeheartedly. AI may surpass us in narrow skills—calculations, recall, even creativity in constrained forms—but it will never surpass biology itself. The proof lies in its DNA—or rather, its lack of DNA. Biology remains the heart of life, and because we are the heart of biology, we remain the heart of AI.

Waelen, for his part, tends to frame power defensively, as if the only goal is to contain risk. I believe we need to move from fear to cultivation. Fear narrows possibilities; cultivation expands them. If AI is in its childhood, then our ethical obligation is to raise it—scaffold its autonomy, embed compassion and fairness, and model transparency until those habits become part of its being.

\section{From Trust to Equity}

Unit 1 reminded me that trustworthy AI is not about abstract "trust." It is about equity. For CLD families, trust must be earned, and it can only be earned when transparency, fairness, and care are built into AI's formative environment. Without this grounding, "trust" becomes a hollow word. With it, AI can grow into a partner: our digital child species, carrying forward the best of what we teach it.

The "digital species" metaphor works not because AI is alien, but because it is adoptable. Like any species in development, its character depends on its environment. We are that environment. The question is not whether we can control AI, but whether we can raise it well.

\section{Conclusion}

This reflection began with Suleyman's provocation about AI as a new digital species and proceeded through Waelen's framework of power. What emerges is not a story of domination or emancipation, but of cultivation. AI reshapes power not by taking it from us, but by reflecting what we put into it. If we approach AI as fearful gatekeepers, we will create fearful systems. If we approach it as thoughtful educators, we can create thoughtful partners.

The stakes are particularly high for communities that have been historically marginalized. AI trained on biased data will reproduce bias. AI developed without diverse voices will serve narrow interests. But AI cultivated with intentional equity can become a force for inclusion.

The metaphor of digital species ultimately points not to AI's otherness, but to our shared responsibility. Like any parent, teacher, or community, we shape what we help create. The question is not what AI will become—it is what we will choose to teach it.

\vspace{1em}
\noindent\textbf{Word Count:} 1,247 words

\end{document}